Machine learning guided batched design\\ of a bacterial Ribosome Binding Site

### 2022-04-20 paper/main.tex
Final paper no highlights.

In summary, 120 out of 450 sequences (BPS-NC, BPS-C, UNI, PPM) are experimental controls created by sequence randomisation.
Only a few of the random sequences were promising, but they were still 20\% weaker than the benchmark sequence's TIR (Figure 3A).
In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (\hl{Note, this applies only to the 7 bps on either side of the core; }see Supplementary Figure S1 and work by \mbox{\textcite{Jeschek2016}}). Sequences from randomization of the core RBS (BPS-C, UNI, PPM) are more appropriate controls, from which the best sequence achieved only about 65\% of the benchmark TIR.
These results show that generating a strong RBS sequence by random mutation is a non-trivial task, when the tested data set is relatively small.
In contrast, our Bandit-driven design gives much better results, with RBSs getting close to benchmark performance and even exceeding it.\\

@@

\hl{We provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
\cite{Schober2018,  Kang2019}.
The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise.
However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
\hl{
This is because, in the Design part, we select RBSs based on the ranks of the TIR predictions instead of numerical predictions.
% in this task, we are maximising the TIR of RBS and so we are looking for RBS with high rank 1, regardless of its numerical TIR.
}
This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
\\

@@

We present some important characteristics of the tested RBSs in Table 1, demonstrating the effectiveness of our Build and Test phases.
Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows the logo generated for all tested sequences).
It is generally understood that guanine-rich sequences promote strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with \hl{higher probability of having a high TIR value}.\\

@@

We discovered an interesting phenomenon of large edit distances between RBSs with high TIRs.
In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR \hl{(we assume here that our data set includes most of the strongest RBSs for each distance - sequence combination)}.
We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
Figure \ref{fig:Library characteristics}B shows the edit distance that is required for positive change in TIR for an RBS with TIR $>0.75$.
For RBSs with high TIRs ($>1$), the minimum distance that is required to increase the TIR is 2, with edit distances between 2 and 5 giving similar results.
For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
\\

This means that as the TIRs of examined RBSs increase, exploring sequences which are increasignly dissimilar to the current candidates tends to give more meaningful improvement.
As long as this does not impact targeted methods like machine learning-guided design, this implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long \hl{(Figure} \ref{fig: Swarmplot and Quantplot}A \hl{shows comparison between efficiency of random versus directed sequence generation)}.
In such cases, a directed process, like the one described here, should be strongly encouraged.
This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than random changes \cite{Jackel2008}.\\

@@

In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
In the Learn phase, we used Gaussian process regression to predict the TIR mean and \hl{uncertainty of that prediction}.
To represent RBSs and capture the similarities between them, we choose to use
the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
In the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
In the Build and Test phase, we performed our experiments using laboratory
automation to increase their speed, reliability and reproducibility.
Using our proposed workflow and testing 450 RBS variants in 4 DBTL cycles,
we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.
In the rest of the section, we first revisit our overall experimental goal and challenges encountered in each phase of the
DBTL cycle (Figure~\ref{fig: Flowchart}).
We then link our proposed methods with related work and discuss the potential generalisation of our framework.
% Then we will discuss the lessons learned for 
We further discuss our design choices and open questions in this line of research.
We refer \cite{zhang_synbio2021} as more detailed discussion.\\

@@

\hl{To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator} \cite{Salis2009} \hl{and used our approach to simulate the search for the strongest designs in limited number of rounds.
We show these results in Supplementary Figure} \ref{fig: Swarmplot and Quantplot Salis}, \hl{but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.}\\

In this study we have limited the number of design rounds to four.
There were a number of reasons for this, including limitations on time and money, but also because the results obtained showed that we have achieved our goal of generating very strong RBS designs.
There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however, this has to be put in the context of limited resources.
For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available \hl{(for example where there is no fluorescent readout available)}, while still providing the required results (optimised part). \\

There are still open questions that need to be addressed for the application of machine learning in synthetic biology.
Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty \hl{quantification}?
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
Similarly, the re-usability of the obtained data set is currently unknown.
The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.
Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
\hl{In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance} \mbox{\cite{Radivojevic2020}}.
%While Radivojevic \textcite{Radivojevic2020} provides a Bayesian ensemble framework, we focus on Gaussian process and provides methods to design batch recommendation better.}
%In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
\\
% \section{Conclusion}

@@

In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments \hl{and so the percentage of the space that is measured will inevitably decrease, which might impact the expected recommendation results}.
To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\

@@

For the regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
%As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
\hl{GPR can predict the posterior mean of the RBS sequences in the design space. At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.}

@@

The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
\hl{
Among various string kernels which have been studied by \mbox{\textcite{Ben-Hur2008}} and we will introduce briefly in Supplementary \mbox{\ref{sec: Choices of Kernels}}, we chose the one with the best reported performance, the \textit{weighted degree kernel with shift} (WDS) \mbox{\cite{ratsch_rase_2005_wds, Ben-Hur2008}},  
to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$.  }
WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
The maximum substring length is specified by $\ell$.
The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
This could be useful when there is a shift between two sequences.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. \\
% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.

@@

$S(l)$ determines the maximum shift at position $l$. 
Furthermore, we normalise the kernel with centring and unit-norm in terms of all sequences in the design space. 
% In our design, the length of RBS sequence $L = 20$.
\hl{
The hyperparameters for kernel, including maximum substring length $\ell = 6$, maximum shift length $S(l) = 1$, and the standard deviation of GP noise $\alpha = 1$ were chosen based on 10-repeat 5-fold cross validation.
The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$. }
% ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).

@@

To recommend the RBS sequences to query in the next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm
\cite{lattimore2020bandit}
% \cite{srinivas2012information}
, which provides  $\textit{exploration-exploitation balance}$.
On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
More precisely, the UCB algorithm selects RBS sequences $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
\begin{align}
\label{Eq: GPUCB}
    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
\end{align}
where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation \hl{(that is uncertainty of the prediction)} at Round $t$ for the sequence $\mathbf{x}_i$.
We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\

@@

\hl{To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver. 2.0 \mbox{\cite{Salis2009}} and evaluated our model using that data.
We normalised the TIR values by dividing them by the TIR of the benchmark sequence, and show the TIR ratio mapping between the RBS calculator and TIRs obtained experimentally for this work in Figure \mbox{\ref{fig:TIR ratio salis micro}}.
We observe that the RBS calculator values are not well aligned to the measured values, which underlines the need for development of new prediction and design methods. 
Among all 4096 RBS sequences, 38 RBS were predicted to have higher TIR than the benchmark sequence by the RBS calculator. We call those sequences the \textit{optimal set}, and the goal is to find as many sequences from the optimal set as permitted by the budget (450).

To do this, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 4 rounds.
We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that our model can identify above 75\% of the optimal set in up to 2 rounds and 99\% of them in up to 3 rounds.
We reported the detailed statistics for the 100 runs in Table \mbox{\ref{table: Important statistics for the 100 synthetic experiments}}.
Figure \mbox{\ref{fig: Swarmplot and Quantplot Salis}} shows plot analogues to these in Figure \mbox{\ref{fig: Swarmplot and Quantplot}}  but for one of the runs using the RBS calculator data.
Our algorithm has successfully found most of the sequences in optimal set in the first and second rounds, thus the recommended TIRs have an apparent drop in the following rounds (as there are no good sequences left to recommend).
This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.}

>>>>

In summary, 120 out of 450 sequences (BPS-NC, BPS-C, UNI, PPM) are experimental controls created by sequence randomisation.
Only a few of the random sequences were promising, but they were still 20\% weaker than the benchmark sequence's TIR (Figure 3A).
In fact, these 80\% TIR ratio sequences were created by randomising the sequence outside of the core RBS region (BPS-NC), which was statistically shown not to be significantly impactful on the TIR (Note, this applies only to the 7 bps on either side of the core; see Supplementary Figure S1 and work by \mbox{\textcite{Jeschek2016}}). Sequences from randomization of the core RBS (BPS-C, UNI, PPM) are more appropriate controls, from which the best sequence achieved only about 65\% of the benchmark TIR.
These results show that generating a strong RBS sequence by random mutation is a non-trivial task, when the tested data set is relatively small.
In contrast, our Bandit-driven design gives much better results, with RBSs getting close to benchmark performance and even exceeding it.\\

@@

We provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient
\cite{Schober2018,  Kang2019}.
The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise.
However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
This is because, in the Design part, we select RBSs based on the ranks of the TIR predictions instead of numerical predictions.
% in this task, we are maximising the TIR of RBS and so we are looking for RBS with high rank 1, regardless of its numerical TIR.
This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
\\

@@

We present some important characteristics of the tested RBSs in Table 1, demonstrating the effectiveness of our Build and Test phases.
Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (Figure \ref{fig: All_logo} shows the logo generated for all tested sequences).
It is generally understood that guanine-rich sequences promote strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandit algorithm's bias towards the G-rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G-rich sequences as the ones with higher probability of having a high TIR value.\\

@@

We discovered an interesting phenomenon of large edit distances between RBSs with high TIRs.
In other words, given an RBS with high TIR, multiple sequence edits need to occur to reach another RBS with high TIR (we assume here that our data set includes most of the strongest RBSs for each distance - sequence combination).
We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
Figure \ref{fig:Library characteristics}B shows the edit distance that is required for positive change in TIR for an RBS with TIR $>0.75$.
For RBSs with high TIRs ($>1$), the minimum distance that is required to increase the TIR is 2, with edit distances between 2 and 5 giving similar results.
For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
\\

This means that as the TIRs of examined RBSs increase, exploring sequences which are increasignly dissimilar to the current candidates tends to give more meaningful improvement.
As long as this does not impact targeted methods like machine learning-guided design, this implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs with a limited budget.
In other words, because the examined sequence is relatively short (6 bp in a wider 20 bp context) the time required to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long (Figure \ref{fig: Swarmplot and Quantplot}A shows comparison between efficiency of random versus directed sequence generation).
In such cases, a directed process, like the one described here, should be strongly encouraged.
This observation is in line with approaches seen in other disciplines, e.g. protein engineering, where more directed changes yield better results than random changes \cite{Jackel2008}.\\

@@

In this work, we show how a machine learning guided approach, and high-throughput, automated laboratory methods can be jointly applied to efficiently optimise a small genetic part, in this case maximising the TIR of a bacterial RBS.
In the Learn phase, we used Gaussian process regression to predict the TIR mean and uncertainty of that prediction.
To represent RBSs and capture the similarities between them, we choose to use
the Weighted Degree Kernel with Shift method, which fits well with Gaussian processes.
In the Design phase we used an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested in batches.
In the Build and Test phase, we performed our experiments using laboratory
automation to increase their speed, reliability and reproducibility.
Using our proposed workflow and testing 450 RBS variants in 4 DBTL cycles,
we designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs in this genetic context by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used as a basis for future studies.
In the rest of the section, we first revisit our overall experimental goal and challenges encountered in each phase of the
DBTL cycle (Figure~\ref{fig: Flowchart}).
We then link our proposed methods with related work and discuss the potential generalisation of our framework.
% Then we will discuss the lessons learned for 
We further discuss our design choices and open questions in this line of research.
We refer \cite{zhang_synbio2021} as more detailed discussion.\\

@@

To further test the performance of our predictions and recommendations we have generated a TIR label for each of the 4096 RBS sequences using the RBS calculator \cite{Salis2009} and used our approach to simulate the search for the strongest designs in limited number of rounds.
We show these results in Supplementary Figure \ref{fig: Swarmplot and Quantplot Salis}, but in short, our approach was able to  identify above 75\% of the sequences stronger than the benchmark in up to 2 rounds and 99\% of them in up to 3 rounds.
This surprisingly high efficiency can be partially attributed to the source of the TIR predictions, which being generated algorithmicaly are easier to predict by our algorithm than real world data.
Nevertheless, it shows that the algorithm can efficiently learn the sequence to TIR relationship and exploit it subsequently in a recommendation task.\\

In this study we have limited the number of design rounds to four.
There were a number of reasons for this, including limitations on time and money, but also because the results obtained showed that we have achieved our goal of generating very strong RBS designs.
There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however, this has to be put in the context of limited resources.
For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
Compared to solutions like the one reported by Hollerer \cite{Hollerer2020}, our solution can be used when a high-volume method for data-generation is not available (for example where there is no fluorescent readout available), while still providing the required results (optimised part). \\

There are still open questions that need to be addressed for the application of machine learning in synthetic biology.
Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty quantification?
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
Similarly, the re-usability of the obtained data set is currently unknown.
The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.
Finally, the practical optimal exploration-exploitation balance between rounds and samples is still an open question.
In our work we have steered the recommendation process towards exploitation as we progressed through the rounds, in line with other works addressing this balance \mbox{\cite{Radivojevic2020}}.
%While Radivojevic \textcite{Radivojevic2020} provides a Bayesian ensemble framework, we focus on Gaussian process and provides methods to design batch recommendation better.}
%In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.
\\
% \section{Conclusion}

@@

In future we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
However, it is important to reiterate that the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments and so the percentage of the space that is measured will inevitably decrease, which might impact the expected recommendation results.
To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\

@@

For the regression model, we have used a Bayesian non-parametric approach called \textit{Gaussian Process Regression (GPR)} \cite{Rasmussen2004,srinivas2012information, romero_navigating_2013}.
We model $f$ as a sample from a \textit{Gaussian Process} $\mathcal{G} \mathcal{P}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x'}))$, which is specified by the mean function $\mu(\mathbf{x})=\mathbb{E}[f(\mathbf{x})]$ and the kernel function (also called \textit{covariance function}) $k\left(\mathbf{x}, \mathbf{x}^{\prime}\right)=\mathbb{E}[(f(\mathbf{x})-\left.\mu(\mathbf{x}))\left(f\left(\mathbf{x}^{\prime}\right)-\mu\left(\mathbf{x}^{\prime}\right)\right)\right]$.
%As illustrated in Figure \ref{fig: Gaussian Process Regression Example.}, 
GPR can predict the posterior mean of the RBS sequences in the design space. At the same time, it can provide the posterior standard deviation, which represents the level of uncertainty for the prediction.

@@

The choice of kernel function is critical for accurate predictions, since it controls smoothness and amplitude of the function to be modelled.
For Bandit designs in Round 0, since we only had access to a limited number of data points from the literature, we chose to use one of the basic string kernels, the \textit{spectrum kernel} \cite{leslie2001spectrum} to process the core 6 bp and dot product kernel \cite{Rasmussen2004} (with one-hot embedding) to process the 7 bp flanking sequences both upstream and downstream of the core sequence.
Among various string kernels which have been studied by \mbox{\textcite{Ben-Hur2008}} and we will introduce briefly in Supplementary \mbox{\ref{sec: Choices of Kernels}}, we chose the one with the best reported performance, the \textit{weighted degree kernel with shift} (WDS) \mbox{\cite{ratsch_rase_2005_wds, Ben-Hur2008}},  
to represent the RBS sequences in subsequent rounds and specify the kernel function of $GP$. 
WDS is a type of a string kernel, which takes two sequences as inputs and outputs a scalar value which represents the similarities between the two sequences.  
WDS kernel does this by counting the matches of substrings of a certain length (i.e. kmers) that constitute the sequence.
The maximum substring length is specified by $\ell$.
The WDS takes into account the positional information by counting substrings starting from different positions, where the start position is specified by $l$.
Additionally, the WDS kernel considers the shifting of substrings, with the maximum shift specified by $s$.
This could be useful when there is a shift between two sequences.
For example, two sequences A\textbf{CCTGA} and \textbf{CCTGA}A are in 1-shift. \\
% there is a common part \textbf{CCTGA} which is begins at the 2nd nucleotide in sequence 1 and at the 1st nucleotide in sequence 2, hence the calculated shift would be 1.

@@

$S(l)$ determines the maximum shift at position $l$. 
Furthermore, we normalise the kernel with centring and unit-norm in terms of all sequences in the design space. 
% In our design, the length of RBS sequence $L = 20$.

The hyperparameters for kernel, including maximum substring length $\ell = 6$, maximum shift length $S(l) = 1$, and the standard deviation of GP noise $\alpha = 1$ were chosen based on 10-repeat 5-fold cross validation.
The values of hyperparameters considered were $\ell \in \{3,4,5,6\}$, $S(l) = \{0,1,2\}$, $\alpha = [0.5, 1, 2]$.
% ($\ell=6$, and $S(l)= 1$, $\alpha = 2$).

@@

To recommend the RBS sequences to query in the next round, we have used the \textit{Upper Confidence Bound (UCB)} batch algorithm
\cite{lattimore2020bandit}
% \cite{srinivas2012information}
, which provides  $\textit{exploration-exploitation balance}$.
On one hand, UCB exploits the function in terms of the design space, that is to pinpoint sequences that are believed to have high labels (i.e. high predicted mean); 
on the other hand, UCB explores the design space where we have little information and sequences have a chance to have high labels (i.e. high predicted standard deviation).
More precisely, the UCB algorithm selects RBS sequences $\mathbf{x}_i \in \mathcal{D}$ with the maximum upper confidence bound at Round $t$, i.e.
\begin{align}
\label{Eq: GPUCB}
    \operatorname{argmax}_{\mathbf{x}_i \in \mathcal{D}} \left( \mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)\right),
\end{align}
where $\beta_t$ is a hyperparameter balancing the exploration and exploitation, 
$\mu_t(\mathbf{x}_i), \sigma_t(\mathbf{x}_i)$ are the predicted mean and standard deviation (that is uncertainty of the prediction) at Round $t$ for the sequence $\mathbf{x}_i$.
We call $\mu_{t-1}(\mathbf{x}_i) + \beta_t \sigma_{t-1}(\mathbf{x}_i)$ the \textit{UCB score} of sequence $\mathbf{x}_i$ at Round $t-1$.\\

@@

To test the reliability of our proposed model, we have generated TIR predictions for all 4096 RBS sequences using the RBS  calculator ver. 2.0 \mbox{\cite{Salis2009}} and evaluated our model using that data.
We normalised the TIR values by dividing them by the TIR of the benchmark sequence, and show the TIR ratio mapping between the RBS calculator and TIRs obtained experimentally for this work in Figure \mbox{\ref{fig:TIR ratio salis micro}}.
We observe that the RBS calculator values are not well aligned to the measured values, which underlines the need for development of new prediction and design methods. 
Among all 4096 RBS sequences, 38 RBS were predicted to have higher TIR than the benchmark sequence by the RBS calculator. We call those sequences the \textit{optimal set}, and the goal is to find as many sequences from the optimal set as permitted by the budget (450).

To do this, we have applied our approach to this data set - first we have chosen 90 sequences at random and after teaching the algorithm using the obtained numbers we have generated 90 further sequences to be tested (using UCB for recommendations) and did this for a total of 4 rounds.
We have repeated this experiment 100 times to increase our confidence in the results, by doing so we have found that our model can identify above 75\% of the optimal set in up to 2 rounds and 99\% of them in up to 3 rounds.
We reported the detailed statistics for the 100 runs in Table \mbox{\ref{table: Important statistics for the 100 synthetic experiments}}.
Figure \mbox{\ref{fig: Swarmplot and Quantplot Salis}} shows plot analogues to these in Figure \mbox{\ref{fig: Swarmplot and Quantplot}}  but for one of the runs using the RBS calculator data.
Our algorithm has successfully found most of the sequences in optimal set in the first and second rounds, thus the recommended TIRs have an apparent drop in the following rounds (as there are no good sequences left to recommend).
This might suggest that it might be valuable to add an additional stopping condition for the DBTL cycle - decreasing round performance.
